---
title: "GBR Spatial Data"
format: 
  html:
    toc: false
    code-fold: true
    self-contained: true
editor: visual
---

The `download_gbr_spatial()` function returns one of four spatial datasets:

```{r}
#| message: false
#| warning: false
#| cache: true
library(dhw)

GBR_base <- download_gbr_spatial(return = "base")
GBR_reefs <- download_gbr_spatial(return = "combined")
GBR_hull <- download_gbr_spatial(return = "hull")
GBR_outline <- download_gbr_spatial(return = "outline")

```

### GBR_base

`return = "base"` returns the raw data downloaded from

`https://nextcloud.eatlas.org.au/s/xQ8neGxxCbgWGSd/download/TS_AIMS_NESP_Torres_Strait_Features_V1b_with_GBR_Features.zip"`

The dataset spans 8390 unique `LABEL_ID` including all GBR features (`FEAT_NAME` includes Bank, Reef, Terrestrial Reef, Rock, Other, Mainland, Cay, Island, Sand)

```{r}

GBR_base
#unique(GBR_base$LABEL_ID) |> length()
#unique(GBR_base$FEAT_NAME)

```


### GBR_reefs

`return = "combined"` groups reefs by `LABEL_ID`, dissolves to one geometry per ID (`summarise(st_union)`), calculates polygon area (`st_area()`) for each `LABEL_ID`, and returns 4612 unique `LABEL_ID` as either polygons or multipolygons.


```{r}

GBR_reefs

#unique(GBR_reefs$LABEL_ID) |> length()
#unique(GBR_base$FEAT_NAME)

```

In the full GBR dataset, "LABEL_ID" is split into multiple subgroups. For example, `LABEL_ID` `09-355` includes 26 sublevels (a,b,c,d, through to a1, a0). This includes `Wapa Reef` by `LOC_NAME_S` but also `U/N Sand Bank`. With `return="combined"` all sublevels of LABEL_ID are combined to a single multipolygon which adopts the `LOC_NAME_S` (see below for `Wapa Reef` example).

Non-GBR content (including Mornington, Forsyth &	South Wellesley Islands in the Gulf of Carpenteria, and all shoals `Y_COORD` < -25) and are additionally removed.

```{r}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 10
library(tidyverse)
library(sf)
library(ggplot2)

GBR_base_09355 <- GBR_base |>
  dplyr::filter(stringr::str_detect(LABEL_ID, "^09-355"))

GBR_reef_09355 <- GBR_reefs |>
  dplyr::filter(stringr::str_detect(LABEL_ID, "^09-355"))


library(patchwork)

a <- ggplot() + theme_bw() + geom_sf(data=GBR_base_09355, aes(fill=LOC_NAME_S)) + ggtitle("return=base") + 
  ggplot2::theme(
  legend.position = "left",
  legend.text  = ggplot2::element_text(size = ggplot2::rel(0.6)),
  legend.title = ggplot2::element_text(size = ggplot2::rel(0.6)),
  legend.key.size = grid::unit(0.25, "lines")
)

b <- ggplot() + theme_bw() + geom_sf(data=GBR_reef_09355, aes(fill=LABEL_ID)) + ggtitle("return=combined")

a|b

```


### GBR_hull

`return = "hull"` returns a single polygon, which is a single convex hull around all `FEAT_NAME` = "Reef", "Terrestrial Reef", "Island", "Rock", "Bank" with an additional 1km buffer. As the original file included a lake in outback Queensland the subset of `FEAT_NAME` that overlap with the `LOC_NAME_S == "Mainland (99-000)")` and non-GBR content (including Mornington, Forsyth &	South Wellesley Islands in the Gulf of Carpenteria, and `Y_COORD` < -25) and are removed prior to creating the convex hull.



```{r}

GBR_hull

#unique(GBR_reefs$LABEL_ID) |> length()
#unique(GBR_base$FEAT_NAME)

```

```{r}

ggplot() + theme_bw() + geom_sf(data=GBR_hull, fill="aquamarine4", alpha=0.2)

```

### GBR_outline


GBR_outline returns the Austrlian mainland via the largest area from the original `GBR_base`:  

```{r}

GBR_outline

```

```{r}

ggplot() + theme_bw() + geom_sf(data=GBR_outline, fill="aquamarine2", alpha=0.2)

```

### Spatial verifications


Check which `LABEL_ID` reefs in `GBR_reefs` overlap with `GBR_reefs`:

```{r}

GBR_reefs_chk <- GBR_reefs |>
  sf::st_make_valid() |>
  sf::st_transform(sf::st_crs(GBR_hull)) |>
  dplyr::mutate(
    within_hull = sf::st_within(geometry, sf::st_geometry(GBR_hull), sparse = FALSE)[, 1],
    intersects_hull = sf::st_intersects(geometry, sf::st_geometry(GBR_hull), sparse = FALSE)[, 1]
  )

GBR_reefs_chk |>
  sf::st_drop_geometry() |>
  dplyr::count(within_hull, intersects_hull)


```




Check with aerial bleaching data for potential missing `LABEL_ID`


- 1998_BerklemansGBRData.csv
- 2002_BerklemansGBRData.csv
- 2016_HughesGBRAerial.csv
- 2017_HughesGBRAerial.csv
- 2020_HughesGBRAerialwArea.csv
- 2022_GBRMPA_GBRAerialData.csv
- 2024GBRReefAvg_AerialBLC_Final_QC_2Apr2024.csv
- 2025GBR_Aerial_ReefAVGBLC_QC_Final.csv


all 1826 `LABEL_ID` from aerial surveys present in `GBR_reefs`:

```{r}

gbr_shape_aerial <- readRDS("/Users/rof011/GBR-dhw/datasets/gbr_shape_aerial.rds")

gbr_shape_aerial |> filter(!LABEL_ID %in% GBR_reefs$LABEL_ID)


```


all 5497 polygons in aerial surveys present in `GBR_reefs`:

```{r}

gbr_shape_aerial <- readRDS("/Users/rof011/GBR-dhw/datasets/gbr_shape_aerial.rds")


GBR_shape_aerial_chk <- gbr_shape_aerial |>
  sf::st_make_valid() |>
  sf::st_transform(sf::st_crs(GBR_hull)) |>
  dplyr::mutate(
    within_hull = sf::st_within(geometry, sf::st_geometry(GBR_hull), sparse = FALSE)[, 1],
    intersects_hull = sf::st_intersects(geometry, sf::st_geometry(GBR_hull), sparse = FALSE)[, 1]
  )

GBR_shape_aerial_chk |>
  sf::st_drop_geometry() |>
  dplyr::count(within_hull, intersects_hull)

```
