---
title: "Validating DHW climatologies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating DHW climatologies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


### Validate climatologies 

The `dhw` package was created to track reproducibility in code while applying the CRW algorithms to additional datasets and further analysis.

To validate the climatology and maximum monthly mean take a spatial subset of the NOAA CoralTemp and the NOAA Daily climatology files for a single pixel (Lizard Island) spanning 1:


```{r, fig.width=9, fig.height=7, message=FALSE}

library(dhw)

plot_mm(input = unwrap(lizard_OISST_raster), lon = 145.25, lat = -14.5)


```

### verify Degree Heating Weeks

The `dhw` package was created to track reproducibility in code while applying the CRW algorithms to additional datasets.

To validate the `create_climatology()` functions, take a spatial subset of the NOAA CoralTemp and the NOAA Daily climatology files for a single pixel (Lizard Island) spanning 1:

(NOAA Climatology available here: [ct5km_climatology_v3.1.nc](https://www.star.nesdis.noaa.gov/pub/sod/mecb/crw/data/5km/v3.1_op/climatology/) - Contains 12 monthly mean SST climatologies for deriving CRW's daily global 5km SST Anomaly product. The maximum pixel-based values from among the 12 monthly mean SST climatologies form the Maximum Monthly Mean (MMM) SST climatology, which is then used to derive CRW's daily global 5km coral bleaching heat stress products)



```{r, message=FALSE}

library(sf)
library(terra)
library(tidyverse)
library(dhw)


grid_point <- data.frame(longitude = 145.405, latitude = -14.655) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  vect() |>
  buffer(width = 0.001)


CoralTempSST <- rast("/Users/rof011/GBR-dhw/datasets/coraltemp/GBR_coraltemp_v3.1_sst.nc") %>% 
  crop(., grid_point) |> 
  project("EPSG:4283")


CoralTempDailyClimatology <- rast("/Users/rof011/GBR-dhw/datasets/coraltemp/ct5km_climatology_v3.1.nc") %>% 
  crop(., grid_point) |> 
  select(2:13) |> 
  project("EPSG:4283")



```

From the CoralTemp data, run `create_climatology()` to generate SST, Daily Climatologies, MM, MMM, SSTA, HS, DHW:

```{r}

test_SST <- create_climatology(CoralTempSST)

```

Comparing the two datasets, June - December mostly match, but MM derived from the `create_climatology()` for Jan-May are (slightly) inconsistent with the NOAA baseline: 

```{r}

data.frame(month=month.abb, 
           dhw_library_mm = values(test_SST$mm) |> as.numeric() |> round(2),
           NOAA_mmm = values(CoralTempDailyClimatology) |> as.numeric()) |> 


ggplot() + theme_bw() +
  geom_point(aes(NOAA_mmm, dhw_library_mm, fill=month), linewidth=0, shape=21, size=2.5, alpha=0.4) +
  stat_smooth(aes(NOAA_mmm, dhw_library_mm), method=lm) +
  coord_fixed()

```


AFAIK this difference is due to the NOAA calculation of the Daily Climatologies, which are based on the full SST Timeseries (1985-01-01 to 2012-12-31), while CoralTemp is only available 1985-06-01 to 2012-12-31.

To test this, we can substitute the CoralTempDailyClimatology back into the climatology calculations. First download the complete NOAA dataset for a subset of time (2015-2017) for all variables (`sst`, `ssta`, `hs`, `dhw`):

```{r, eval=FALSE}

library(rerddap)

NOAA_CRW_a <- griddap(
    datasetx = 'NOAA_DHW',
    time = c("2015-01-01", "2018-01-01"),
    latitude = c(-14.655, -14.655),
    longitude = c(145.405, 145.405),
    #fields = c('CRW_SST', 'CRW_SSTANOMALY', 'CRW_HOTSPOT', 'CRW_DHW'),
    fmt = "csv"
  )

NOAA_CRW_b <- griddap(
    datasetx = 'NOAA_DHW',
    time = c("2018-01-01", "2021-01-01"),
    latitude = c(-14.655, -14.655),
    longitude = c(145.405, 145.405),
    #fields = c('CRW_SST', 'CRW_SSTANOMALY', 'CRW_HOTSPOT', 'CRW_DHW'),
    fmt = "csv"
  )

NOAA_CRW_c <- griddap(
    datasetx = 'NOAA_DHW',
    time = c("2021-01-01", "2024-05-15"),
    latitude = c(-14.655, -14.655),
    longitude = c(145.405, 145.405),
    #fields = c('CRW_SST', 'CRW_SSTANOMALY', 'CRW_HOTSPOT', 'CRW_DHW'),
    fmt = "csv"
  )

NOAA_CRW <- rbind(NOAA_CRW_a, NOAA_CRW_b, NOAA_CRW_c)

#usethis::use_data(NOAA_CRW, overwrite=TRUE)


```


To compare this subset, run the individual components of the `create_climatology()` except replace the `calculated_mm` with the NOAA CRW `CoralTempDailyClimatology` amd visualise the data:


```{r, fig.width=9, fig.height=9}



calculated_mm <- calculate_monthly_mean(CoralTempSST)
calculated_mmm <- calculate_maximum_monthly_mean(mm = CoralTempDailyClimatology)
calculated_dc <- calculate_daily_climatology(sst_file = CoralTempSST, mm = CoralTempDailyClimatology)
calculated_anomalies <- calculate_anomalies(sst_file = CoralTempSST, climatology = calculated_dc)
calculated_hotspots <- calculate_hotspots(mmm = calculated_mmm, sst_file = CoralTempSST)
calculated_dhw <- calculate_dhw(hotspots = calculated_hotspots)
calculated_baa <- calculate_baa(hotspots = calculated_hotspots, dhw = calculated_dhw)


# convert to df
calculated_sst <- CoralTempSST |> as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> rename(calculated_SST=values) |> 
  filter(time >= as.Date("2015-01-01")) |>  filter(time < as.Date("2024-05-15")) |> select(-x, -y, -layer)
calculated_dc <- calculated_dc |> as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> rename(calculated_DC=values) |> 
  filter(time >= as.Date("2015-01-01")) |>  filter(time < as.Date("2024-05-15")) |> select(-x, -y, -layer)
calculated_anomalies <- calculated_anomalies |> as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> rename(calculated_SSTA=values) |> 
  filter(time >= as.Date("2015-01-01")) |>  filter(time < as.Date("2024-05-15")) |> select(-x, -y, -layer)
calculated_hotspots <- calculated_hotspots |> as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> rename(calculated_HS=values) |> 
  filter(time >= as.Date("2015-01-01")) |>  filter(time < as.Date("2024-05-15")) |> select(-x, -y, -layer)
calculated_dhw <- calculated_dhw |> as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> rename(calculated_DHW=values) |> 
  filter(time >= as.Date("2015-01-01")) |>  filter(time < as.Date("2024-05-15")) |> select(-x, -y, -layer)
calculated_baa <- calculated_baa |> as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> rename(calculated_BAA=values) |> 
  filter(time >= as.Date("2015-01-01")) |>  filter(time < as.Date("2024-05-15")) |> select(-x, -y, -layer)


daily_climatology_plot <- ggplot() + theme_bw() + 
  geom_line(data=calculated_sst, aes(time,calculated_SST), color="grey") +
  geom_line(data = calculated_dc, aes(time, calculated_DC), color="blue") +
  geom_hline(yintercept=values(calculated_mmm) |> as.numeric(), color="red")  +
  ylab("SST")

library(patchwork)

daily_anomalies_plot <- ggplot() + theme_bw() + 
  geom_line(data=calculated_anomalies, aes(time, calculated_SSTA)) +
  ylab("SST Anomalies")

daily_hs_plot <- ggplot() + theme_bw() + 
  geom_line(data = calculated_hotspots, aes(time, calculated_HS)) +
  ylab("Hotspots")

daily_dhw_plot <- ggplot() + theme_bw() + 
  geom_line(data=calculated_dhw, aes(time, calculated_DHW)) +
  ylab("Degree Heating Weeks")


daily_baa_plot <- ggplot() + theme_bw() + 
  geom_line(data=calculated_baa, aes(time, calculated_BAA)) +
  ylab("Bleaching Alert Area")

(daily_climatology_plot / daily_anomalies_plot / daily_hs_plot / daily_dhw_plot / daily_baa_plot)

```
Third, compare the calculated climatology values for `sst`, `ssta`, `hs`, `dhw` with the original downloaded NOAA `.nc` file to validate the method:


```{r, fig.width=9, fig.height=6}

NOAA_data <- NOAA_CRW_data |> dplyr::select(time, CRW_SST, CRW_SSTANOMALY, CRW_HOTSPOT, CRW_DHW, CRW_BAA) |> 
  mutate(time=as.Date(time)) |> 
  rename(NOAA_SST = CRW_SST,
         NOAA_SSTA = CRW_SSTANOMALY,
         NOAA_HS = CRW_HOTSPOT,
         NOAA_DHW = CRW_DHW,
         NOAA_BAA = CRW_BAA) |> 
  mutate(NOAA_HS = ifelse(NOAA_HS < 1, 0, NOAA_HS)) #|> 
  #mutate(NOAA_DHW = ifelse(NOAA_DHW == 0, NA, NOAA_DHW))
         

# DHW
calculated_data <- 
  left_join(calculated_sst, calculated_anomalies, by = join_by(time)) %>% 
  left_join(., calculated_hotspots, by = join_by(time)) %>% 
  left_join(., calculated_dhw, by = join_by(time)) %>% 
  left_join(., calculated_baa, by = join_by(time)) |> 
  filter(time %in% NOAA_data$time) |> 
  select(time, calculated_SST, calculated_SSTA, calculated_HS, calculated_DHW, calculated_BAA)

combined_data <- left_join(calculated_data, NOAA_data, by = join_by(time))


#View(combined_data |> select(time, NOAA_HS, NOAA_DHW, NOAA_BAA, round(calculated_HS,2), round(calculated_DHW,2), calculated_BAA))

## reextract mm

mm_comparison <- data.frame(month=month.abb, 
           calculated_MM = values(calculated_mm) |> as.numeric() |> round(2),
           NOAA_MM = values(CoralTempDailyClimatology) |> as.numeric())



###

calibrated_plot_SST <- ggplot() + theme_bw() +
  geom_point(data=combined_data, aes(NOAA_SST, calculated_SST), color="indianred1") +
  geom_abline()

calibrated_plot_SSTA <- ggplot() + theme_bw() +
  geom_point(data=combined_data, aes(NOAA_SSTA, calculated_SSTA), color="khaki3") +
  geom_abline()


calibrated_plot_HS <- ggplot() + theme_bw() +
  geom_point(data=combined_data, aes(NOAA_HS, calculated_HS), color="turquoise") +
  geom_abline()

calibrated_plot_DHW <- ggplot() + theme_bw() +
  geom_point(data=combined_data, aes(NOAA_DHW, calculated_DHW), color="lightseagreen") +
  geom_abline()


calibrated_plot_BAA <- ggplot() + theme_bw() +
  geom_point(data=combined_data, aes(NOAA_BAA, calculated_BAA), color="purple") +
  geom_abline() +
  coord_equal()



(calibrated_plot_SST + calibrated_plot_SSTA + calibrated_plot_HS + calibrated_plot_DHW + calibrated_plot_BAA) + plot_layout(nrow=2)
   
```


# 
# ### multiple gridcells
# 
# 
# ```{r, fig.width=9, fig.height=4}
# 
# 
# # Load Coraltemp
# 
# grid_point <- data.frame(longitude = 145.405, latitude = -14.655) |>
#   st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
#   vect() |>
#   buffer(width = 50000)
# 
# CoralTempSST_subset <- rast("/Users/rof011/GBR-dhw/datasets/coraltemp/GBR_coraltemp_v3.1_sst.nc") |>
#   select(2:13) |> 
#   crop(grid_area) |> 
#    (\(x) { crs(x) <- "EPSG:4283"; x })()
# 
# 
# # Load Climatology
# 
# CoralTempDailyClimatology <- rast("/Users/rof011/GBR-dhw/datasets/coraltemp/ct5km_climatology_v3.1.nc") |> 
#   select(2:13) |> 
#   crop(grid_area) |> 
#    (\(x) { crs(x) <- "EPSG:4283"; x })()
# 
# CoralTempDailyClimatology_subset <- CoralTempDailyClimatology
# 
# # mask to match sst
# CoralTempDailyClimatology_subset <- terra::mask(CoralTempDailyClimatology, CoralTempSST_subset)
# 
# # Calculate Climatology
# 
# calculated_mm <- calculate_monthly_mean(sst_file = CoralTempSST_subset)
# calculated_mmm <- calculate_maximum_monthly_mean(mm = CoralTempDailyClimatology)
# calculated_dc <- calculate_daily_climatology(sst_file = CoralTempSST_subset, mm = CoralTempDailyClimatology)
# calculated_anomalies <- calculate_anomalies(sst_file = CoralTempSST_subset, climatology = calculated_dc)
# calculated_hotspots <- calculate_hotspots(mmm = calculated_mmm, sst_file = CoralTempSST_subset)
# calculated_dhw <- calculate_dhw(hotspots = calculated_hotspots)
# calculated_baa <- calculate_baa(hotspots = calculated_hotspots, dhw = calculated_dhw)
# 
# 
# 
# 
# a1 <- ggplot() + theme_bw() + 
#   ggtitle("Jan MM (dhw library)") +
#   tidyterra::geom_spatraster(data=calculated_mm["mm-Jan"]) +
#   scale_fill_distiller(palette="RdBu", limits=c(28.4, 28.6))
# 
# b1 <- ggplot() + theme_bw() + facet_wrap( ~ lyr) +
#   ggtitle("Jan MM (NOAA Raw)") +
#   tidyterra::geom_spatraster(data=CoralTempDailyClimatology["sst_clim_january"]) +
#   scale_fill_distiller(palette="RdBu", limits=c(28.4, 28.6))
# 
# a1|b1
# 
# 
# 
# 
# calculated_anom_max <- terra::app(calculated_dhw, fun = function(x) {
#   return(base::max(x, na.rm = TRUE))
# })
# 
# 
# calculated_hotspots_max <- terra::app(calculated_hotspots, fun = function(x) {
#   return(base::max(x, na.rm = TRUE))
# })
# 
# calculated_dhw_max <- terra::app(calculated_dhw, fun = function(x) {
#   return(base::max(x, na.rm = TRUE))
# })
# 
# calculated_baa_max <- terra::app(calculated_baa, fun = function(x) {
#   return(base::max(x, na.rm = TRUE))
# })
# 
# 
# ```
# 
# 
# 
# 
# 
# ### Timeseries end dates
# 
# 
# ```{r, fig.width=9, fig.height=7}
# 
# 
# # Load Coraltemp
# 
# grid_point <- data.frame(longitude = 145.405, latitude = -14.655) |>
#   st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
#   vect() |>
#   buffer(width = 100)
# 
# CoralTempSST_subset <- rast("/Users/rof011/GBR-dhw/datasets/coraltemp/GBR_coraltemp_v3.1_sst.nc") |> 
#   crop(grid_point) |> 
#    (\(x) { crs(x) <- "EPSG:4283"; x })()
# 
# CoralTempSST_subset[[time(CoralTempSST_subset) < as.Date("1991-01-01")]]
# 
# 
# # Load Climatology
# 
# CoralTempDailyClimatology <- rast("/Users/rof011/GBR-dhw/datasets/coraltemp/ct5km_climatology_v3.1.nc") |> 
#   select(2:13) |> 
#   crop(grid_area) |> 
#    (\(x) { crs(x) <- "EPSG:4283"; x })()
# 
# 
# 
# # Calculate Climatology
# 
# calculated_mm <- calculate_monthly_mean(sst_file = CoralTempSST_subset)
# calculated_mmm <- calculate_maximum_monthly_mean(mm = CoralTempDailyClimatology)
# calculated_dc <- calculate_daily_climatology(sst_file = CoralTempSST_subset, mm = CoralTempDailyClimatology)
# calculated_anomalies <- calculate_anomalies(sst_file = CoralTempSST_subset, climatology = calculated_dc)
# calculated_hotspots <- calculate_hotspots(mmm = calculated_mmm, sst_file = CoralTempSST_subset)
# calculated_dhw <- calculate_dhw(hotspots = calculated_hotspots)
# calculated_baa <- calculate_baa(hotspots = calculated_hotspots, dhw = calculated_dhw)
# 
# 
# 
# NOAA_CRW_subset <- NOAA_CRW |> 
#   as.data.frame() |> 
#   mutate(time = as.Date(time)) |> 
#   select(time, crw_dhw)
# 
# calculated_dhw_subset <- calculated_dhw |> 
#   as.data.frame(xy=TRUE, wide=FALSE, time=TRUE) |> 
#   rename(dhw=values) |> 
#   filter(time %in% NOAA_CRW_subset$time) |> 
#   select(time, dhw)
# 
# 
# combined_subsets <- left_join(NOAA_CRW_subset, calculated_dhw_subset, by = join_by(time))
# combined_subsets_long <- combined_subsets |> 
#   pivot_longer(-time, names_to = "data_source", values_to = "dhw")
# 
# library(patchwork)
# 
# a <- ggplot() + theme_bw() + facet_wrap( ~ data_source) +
#   geom_line(data=combined_subsets_long, aes(time, dhw, color=data_source), show.legend=FALSE)
# 
# b <- ggplot() + theme_bw() +
#   geom_line(data=combined_subsets_long, aes(time, dhw, color=data_source))
# 
# 
# c <- ggplot() + theme_bw() +
#   geom_point(data=combined_subsets, aes(crw_dhw, dhw))
# 
# 
# a/(b|c)
# 

```


